\sysfull

### 2018-04-15 paper/implementation.tex
little corrections in the implementation
Section 6 enriched

We proceed with the explanation of the implementation of \sys programming model. \sys implementation is light and adds very little on top of a plain C language.

\subsection{\sys Application Programming Interface}
\label{sec:coala_api}

@@

\subsection{\sys Kernel Functions Implementation}

@@

\noindent \textbf{New Protected Variables.} The API method \texttt{COALA\_PV} statically allocates (and aligns in memory) a non-volatile variable, whose address will be used by \texttt{RP} and \texttt{WP} API methods to locate the most recent location of the variable itself, which could be either the location of the variable in volatile memory or its location in non-volatile memory.

\noindent \textbf{Task Transitions.} By calling \texttt{COALA\_NEXT\_TASK}, the programmer tells \sys's task manager which task to execute next. The implementation of \sys is built on the assumption that the currently executing task is not interrupted, and only once it completes the next task is run.

\noindent \textbf{Protected Reads and Writes.} When \texttt{RP} and \texttt{WP} API methods are used, \sys searches the volatile memory for the page the protected variable belongs to. If such page is not found in RAM, then a page fault occurs, and the page is fetched from the non-volatile memory. The two functions return a handle to the variable's volatile memory location. Unlike \texttt{RP}, \texttt{WP} marks the page being written as dirty, so that it can be copied to its non-volatile location during the commit phase.

\noindent \textbf{Initialization and Origin Task.} The behavior of the API method \texttt{COALA\_INIT} is very similar to \texttt{COALA\_NEXT\_TASK}, with the addition that all the preliminary kernel initializations are performed inside this function. Moreover, the origin task is scheduled only upon the first boot.

@@

\noindent \textbf{Execution.} The application control is passed to \sys's execution manager by calling \texttt{COALA\_RUN} at the end of the main function. A high-level state machine of the execution manager is depicted in Figure~\ref{fig:coala_state_machine} \todo{Draw a figure reflecting the following description}{Sinan, Carlo}. Upon reboot, the coalescing target is initialized according to the coalescing strategy. Then, the second phase of the commit is resumed, in case the last power failure occurred during this operation. The shadow buffer list is cleared, and the metadata of the volatile buffer is initialized. The pointer to the last coalescing task is recalled from non-volatile memory. Before running, Coala checks whether there is any partial execution to resume, in which case the volatile state, including the PC, is restored. Otherwise, coalescing starts: the current budget is replenished; tasks are run, the history is increased and the current budget is decreased; when the latter is zero, coalescing stops, the pointer to the next task is saved and the commit phase starts. Coalescing resumes after commit phase two ends and the target is updated based on the coalescing strategy. \todo{Go over keywords (coalescing target, current budget, commit phases etc.) and make them consistent with the rest of the paper}{Amjad, Carlo}

\subsection{\sys Paging Implementation}

\noindent \textbf{Memory Virtualization.} The memory virtualization is implemented using three buffers: (i) working (volatile) buffer, (ii) store (non-volatile) buffer and (iii) shadow (non-volatile) buffers. \todo{More explanation required here}{Carlo}

\noindent \textbf{Page Allocation and Page Tags.} When the \sys application accesses a protected variable with either \texttt{RP} or \texttt{WP}, \sys has to search for the variable in RAM first, and then in a non-volatile memory if it not found in RAM. In reality, the \sys kernel searches for pages, not single variables. A page search can be very fast if each page is identified by a unique tag, and if every variable belonging to a page bears the tag in its address. Thus, retrieving a page tag from a protected variable's address boils down to extracting the upper bits of the address itself. The remaining lower bits denote the variable's offset in its page. The complete process is depicted in Figure~\ref{figure:coala_page_tags}.

@@

We note that in order to obtain unique page tags, some alignment requirements have to be met. First, the page size $S$ has to be the power of two. Moreover, pages in non-volatile memory have to be aligned by $S$. Note that, by aligning pages by $2^n$, the alignment is also valid for all sizes $2^x$, where $x \leq n$.

\noindent \textbf{DMA Page Fault and Page Swap.} When a page is not found in RAM, it is fetched from FRAM. \todo{More explanation required here}{Carlo}

\subsection{\sys Scheduling and Committing}

\noindent \textbf{Target Budget Initialization.} Upon a reboot, the target budget (or target coalesced task size) has to be updated depending on the coalescing strategy. \todo{expand text}{Carlo}

\noindent \textbf{Coalescing State Recall.} The last (coalesced) task has to be recalled. More specifically, the function pointer of the last coalesced task has to be retrieved from a non-volatile memory, where is had been saved during the last successful commit. \todo{expand text}{Carlo}

\noindent \textbf{Partial Execution Restoration.} In case there is a valid partial execution checkpoint, the program has to resume from there \todo{expand text}{Carlo}

\noindent \textbf{Coalesced Scheduling.} \todo{expand text}{Carlo}

\noindent \textbf{Intermediate Commit and Ultimate Commit.} When a coalesced task is successfully run to completion, all the dirty pages in SRAM have to be persistently saved to a non-volatile memory. Since last version, no major changes were made to the intermediate commit (SRAM to shadow buffer), but an optimization was applied to the ultimate commit (shadow buffer to store buffer). Store buffer and shadow buffer used to be physically separated in memory. Now, the role of a block of memory (i.e. a page) is interchangeable, meaning that, during the ultimate commit, the shadow buffer pages that were written during the previous commit phase become store buffer pages, and their relative store buffer pages become shadow buffer pages. A page’s role is retrieved using an indirection table, i.e. an array that holds a zero for all the pages whose store buffer resides in the lower memory block, and a one for all the pages whose store buffer is located in the upper memory block. This allows for a lighter commit process, since fewer page moves are performed.

@@

\noindent \textbf{Target Budget Update.} Before starting a new coalesced task, the target budget (or target coalesced task size) has to be updated depending on the coalescing strategy. \todo{expand text}{Carlo}

\todo{API of the partial commit needs to be explained as well}{Carlo}

>>>>

We proceed with the explanation of the implementation of \sys programming model, which is light and whose API adds very little on top of a plain C language.

\subsection{Application Programming Interface}
\label{sec:coala_api}

@@

\subsection{Kernel Functions Implementation}

@@

\textbf{New Protected Variables.} The API method \texttt{COALA\_PV} statically allocates (and aligns in memory) a non-volatile variable, whose address will be used by \texttt{RP} and \texttt{WP} API methods to locate the most recent location of the variable itself, which could be either the location of the variable in volatile memory or its location in non-volatile memory.

\textbf{Task Transitions.} By calling \texttt{COALA\_NEXT\_TASK}, the programmer tells \sys's task manager which task to execute next. The implementation of \sys is built on the assumption that the currently executing task is not interrupted, and only once it completes the next task is run.

\textbf{Protected Reads and Writes.} When \texttt{RP} and \texttt{WP} API methods are used, \sys searches the volatile memory for the page the protected variable belongs to. If such page is not found in RAM, then a page fault occurs, and the page is fetched from the non-volatile memory. The two functions return a handle to the variable's volatile memory location. Unlike \texttt{RP}, \texttt{WP} marks the page being written as dirty, so that it can be copied to its non-volatile location during the commit phase.

\textbf{Initialization and Origin Task.} The behavior of the API method \texttt{COALA\_INIT} is very similar to \texttt{COALA\_NEXT\_TASK}, with the addition that all the preliminary kernel initializations are performed inside this function. Moreover, the origin task is scheduled only upon the first boot.

@@

\textbf{Execution.} The application control is passed to \sys's execution manager by calling \texttt{COALA\_RUN} at the end of the main function. A high-level state machine of the execution manager is depicted in Figure~\ref{fig:coala_state_machine} \todo{Draw a figure reflecting the following description}{Sinan, Carlo}. Upon reboot, the coalescing target is initialized according to the coalescing strategy. Then, the ultimate commit is resumed, in case the last power failure occurred during this operation. The shadow buffer list is cleared, and the metadata of the volatile buffer is initialized. The pointer to the last coalescing task is recalled from non-volatile memory. Before running, Coala checks whether there is any partial execution to resume, in which case the volatile state, including the PC, is restored. Otherwise, coalescing starts: the current budget is replenished; tasks are run, the history is increased and the current budget is decreased; when the latter is zero, coalescing stops, the pointer to the next task is saved and the (intermediate) commit starts. Coalescing resumes after the ultimate commit is complete and the target is updated based on the coalescing strategy. \todo{Go over keywords (coalescing target, current budget, commit phases etc.) and make them consistent with the rest of the paper}{Amjad, Carlo}

\subsection{Paging Implementation}

\sys's memory virtualization is implemented using three buffers: (i) working (volatile) buffer, (ii) store (non-volatile) buffer and (iii) shadow (non-volatile) buffer. The store buffer holds the latest safe copy of all the pages, and it is read-only up to, and including, the ultimate commit. Upon reboot, the working buffer is empty, and it gets filled with pages from non-volatile memory on demand. All reads and writes to protected variables happen in volatile memory. The shadow buffer serves as an intermediate non-volatile buffer where dirty, volatile pages are copied during the intermediate commit. Moreover, it contains pages evicted from SRAM when a \emph{full page fault} occurs.

\textbf{Page Allocation and Page Tags.} When a \sys application accesses a protected variable with either \texttt{RP} or \texttt{WP}, \sys has to search for the variable in RAM first, and then, if not found, in non-volatile memory. In reality, the page manager searches for pages, not single variables. A page search can be very fast if each page is identified by a unique tag, and if every variable belonging to a page bears the tag in its address. Thus, retrieving a page tag from a protected variable's address boils down to extracting the upper bits of the address itself. The remaining lower bits denote the variable's offset in its page. The complete process is depicted in Figure~\ref{figure:coala_page_tags} \todo{this figure doesn't really show page tags and offsets, but paging in general}.
We note that, in order to obtain unique page tags, some alignment requirements have to be met. First, the page size $S$ has to be the power of two. Moreover, pages in non-volatile memory have to be aligned by $S$. Note that, by aligning pages by $2^n$, the alignment is also valid for all sizes $2^x$, where $x \leq n$.

@@

\textbf{Page Fault and Page Swap.} When a page is not found in RAM, a \emph{page fault} it has to be fetched from FRAM. In case that page was in the working buffer and it had been evicted, its most recent non-volatile copy resides in the shadow buffer. For this reason, \sys searches this buffer first. If the requested page is not found in the shadow buffer, then it will be pulled from its store location. To make room for the requested page in SRAM, a volatile page might have to be evicted. If the victim page is dirty, we say that a \emph{full page fault} has occurred, and the victim page has to be copied to the shadow buffer. Page moves are accelerated by the MCU's direct memory access (DMA) module.

% It is important to mention that store and shadow buffers are virtual. By default, \sys allocates two memory banks for pages, $B0$ and $B1$, and all the pages have their store copy in $B0$. A dirty page that during the intermediate commit, or due to eviction, is copied to its shadow location in $B1$, will end up having its store location in $B1$ after the ultimate commit. Thus, during the ultimate commit

\subsection{Coalescing Loop and Commit Phase}

As described above, the main coalescing loop alternates between executing tasks and saving progresses. Static tasks are coalesced as long as the current budget is not null, after which the two commit phases are triggered, the coalescing target is updated and the current budget is replenished.

\textbf{Intermediate Commit and Ultimate Commit.} The reason why the commit process is split in two phases is two-fold: ensure consistency by using atomic operations, and preserving performance by enabling an incremental commit mechanism. When a coalesced task is successfully run to completion, all the dirty pages in SRAM have to be persistently saved to the non-volatile shadow buffer. This is what we call \emph{intermediate commit}: it has to be completed in one shot, and if power fails during this phase the application will have to resume from the beginning of the last coalesced task. When the intermediate commit is done, a flag can be atomically raised to signify that a valid, persistent copy of dirty pages is available in non-volatile memory. From this point on, the commit process can be resumed even if power fails.
To understand the last phase, referred to as \emph{ultimate commit}, it has to be mentioned that store and shadow buffers are virtual. By default, \sys allocates two memory banks for non-volatile pages, $M_0$ and $M_1$, and all the pages have their initial store copy in $M_0$. A dirty page that during the intermediate commit, or due to eviction, is copied to its shadow location in $M_x$, will end up having its store location in $M_x$ after the ultimate commit. In other words, during the ultimate commit dirty pages are not physically moved, instead their store and shadow locations are swapped. This trick makes \sys save some energy during this phase, avoiding redundant page moves. As for page swaps, the intermediate commit makes use of the on-board DMA.

\subsection{Partial execution}

@@
